---
title: 'Health and Economic Impact of Severe Weather (US 1950 - 2011)'
author: "Randy Armknecht"
date: "Saturday, July 18, 2015"
output: html_document
keep_md: true
---

## Synopis
TBD

## Data Processing
### Preparing the environment
It's good practice to set the working directory at the start of a project:
```{r}
setwd("C:/Users/Randy/Documents/Github/Reproducible-Research/storms/")
```

### Loading and Processing the Data

First obtain the data:
```{r}
# Only need to do this once, commenting out for knitr
#data.url <- 'https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2'
#download.file(data.url, destfile="storms.csv.bz2")
```

Now we can load the data we've downloaded:
```{r, cache=TRUE}
data.raw <- read.csv("storms.csv.bz2")
```

And consider some basic information about it:
```{r}
dim(data.raw)
summary(data.raw)
head(data.raw)
```

Create a dataset that is focused on health information:
```{r}
library(dplyr)
data.slim.health <- select(data.raw, BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES)

health <- group_by(data.slim.health, EVTYPE)
health.sum <- summarise(health, 
    count = n(),
    fatalities = sum(FATALITIES),
    injuries = sum(INJURIES),
    f_or_i = fatalities + injuries,
    i_per_event = injuries / count,
    f_per_event = fatalities / count,
    if_per_event = (injuries+fatalities)/count)
```

Create a dataset that is focused on economic information:
```{r}
data.slim.econ <- select(data.raw, BGN_DATE, STATE, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

econ <- group_by(data.slim.econ, EVTYPE)
econ <- filter(econ, PROPDMG > 0)
econ <- mutate(econ, pmulti = gsub("K", 1000, PROPDMGEXP)) 
econ <- mutate(econ, pmulti = gsub("M", 1000000, pmulti))
econ <- mutate(econ, pmulti = gsub("B", 1000000000, pmulti))
econ <- mutate(econ, pmulti = as.factor(pmulti))

conly <- filter(econ, CROPDMG > 0 || PROPDMG > 0)
conly
conly <- mutate(conly, cm = ifelse(CROPDMGEXP == "K", 1000, 
                            ifelse(CROPDMGEXP == "M", 1000000, 
                            ifelse(CROPDMGEXP == "B", 1000000000, NA))),
                       pm = ifelse(PROPDMGEXP == "K", 1000, 
                            ifelse(PROPDMGEXP == "M", 1000000, 
                            ifelse(PROPDMGEXP == "B", 1000000000, NA))))
econ$pmulti
econ$property
for (i in 1:nrow(data.slim.econ)) {
  if(data.slim.econ[i,]$PROPDMG > 0){
    if(data.slim.econ[i,]$PROPDMGEXP == "K"){
      data.slim.econ[i,]$property <- data.slim.econ[i,]$PROPDMG * 1000
    } else if (data.slim.econ[i,]$PROPDMGEXP == "M") {
      data.slim.econ[i,]$property <- data.slim.econ[i,]$PROPDMG * 1000000
    } else if (data.slim.econ[i,]$PROPDMGEXP == "B") {
      data.slim.econ[i,]$property <- data.slim.econ[i,]$PROPDMG * 1000000000
    } else {
      data.slim.econ[i,]$property <- NA
    }
  } else {
    data.slim.econ[i,]$property <- data.slim.econ[i,]$PROPDMGEXP
  }
}

unique(data.slim.econ$PROPDMGEXP)
  
econ <- group_by(data.slim.econ, EVTYPE)
econ.sum <- summarize(econ, count = n())

```

## Results
### Across the United States, which types of events are most harmful with respect to population health?

We can see that Hurricane/Typhoon and Extreme Heat have traditionally caused the greatest impact on population health on a per event basis.
```{r}
tmp <- arrange(filter(health.sum, count > 10), desc(if_per_event))[1:10,]
par(las=2) # Make Label Perpendicular to Axis
par(mar=c(7,10,4,2)) # Adjust Y Margin
barplot(tmp$if_per_event, 
        names.arg=tmp$EVTYPE, 
        horiz=TRUE, 
        cex.names=0.6, 
        cex.axis=0.8,
        xlim=c(0,20),
        col="Dark Red",
        main="Injuries or Fatalities Per Event")
```

In terms of total impact, Tornados have had the greatest affect on population health. The below chart is logrithmic due to the volume of people impacted by Tornados.
```{r}
tmp <- arrange(filter(health.sum, count > 10), desc(f_or_i))[1:10,]
par(las=2) # Make Label Perpendicular to Axis
par(mar=c(7,10,4,2)) # Adjust Y Margin
barplot(tmp$f_or_i, 
        names.arg=tmp$EVTYPE, 
        horiz=TRUE, 
        cex.names=0.6, 
        cex.axis=0.8,
        log="x",
        xlim=c(1000,100000),
        col="Dark Red", 
        main="Total Injuries or Fatalities By Event Type")
```

### Across the United States, which types of events have the greatest economic impact?


## Data Exploration
Consider various information about the event types impact on population health, per event:
```{r}
# Events with the greatest number of injuries per event
arrange(health.sum, desc(i_per_event))[1:10,]

# Events with the greatest number of fatalities per event
arrange(health.sum, desc(f_per_event))[1:10,]

# Again, but only Event Types that have occured at least 10 times
arrange(filter(health.sum, count > 10), desc(i_per_event))[1:10,]
arrange(filter(health.sum, count > 10), desc(f_per_event))[1:10,]

# Let's consider both fatalities and injuries, as both impact population health
arrange(filter(health.sum, count > 10), desc(if_per_event))[1:10,]
```

Consider various information about the event types impact on population health, in total:
```{r}
# Events with the greatest number of injuries
arrange(health.sum, desc(injuries))[1:10,]

# Events with the greatest number of fatalities per event
arrange(health.sum, desc(fatalities))[1:10,]

# Let's consider both fatalities and injuries, as both impact population health
arrange(filter(health.sum, count > 10), desc(f_or_i))[1:10,]
```

Let's take a look at the various states:
```{r}
library(statebins)
states <- group_by(data.slim.health, STATE)
states.sum <- summarise(states, 
    count = n(),
    fatalities = sum(FATALITIES),
    injuries = sum(INJURIES),
    f_or_i = fatalities + injuries)

abvs <- c("AK", "AL", "AR", "AZ", "CA", "CO", "CT",
          "DC", "DE", "FL", "GA", "HI", "IA", "ID",
          "IL", "IN", "KS", "KY", "LA", "MA", "MD",
          "ME", "MI", "MN", "MO", "MS", "MT", "NC",
          "ND", "NE", "NH", "NJ", "NM", "NV", "NY",
          "OH", "OK", "OR", "PA", "RI", "SC", "SD", 
          "TN", "TX", "UT", "VA", "VT", "WA", "WI",
          "WV", "WY")
states.pure <- states.sum[states.sum$STATE %in% abvs,]
states.pure$state <- as.character(states.pure$STATE)
statebins_continuous(as.data.frame(states.pure), brewer_pal="Purples",
                     state_col="state", value_col="f_or_i")


```

Surprisingly Texas takes the worst of it.
```{r}
arrange(states.pure, desc(f_or_i))[1:10,]
```
