---
title: 'Health and Economic Impact of Severe Weather (US 1950 - 2011)'
author: "Randy Armknecht"
date: "Friday, July 24, 2015"
output: html_document
keep_md: true
---

## Synopis
Over the last 60 years more than $419 Billion in economic damage has occured, the bulk by flooding and hurricanes. More than 154,000 people have been either injured or killed by storms.  Given the broad timeframe, and the population of the United States, this number was much lower than anticipated.  What follows is the data processing to setup the analysis, some core results to answer the main questions, and then exploratory analysis as a bonus.

## Data Processing
### Preparing the environment
It's good practice to set the working directory at the start of a project:
```{r}
setwd("C:/Users/Randy/Documents/Github/Reproducible-Research/storms/")
```

### Loading and Processing the Data

First obtain the data:
```{r}
# Only need to do this once, commenting out for knitr
#data.url <- 'https://d396qusza40orc.cloudfront.net/repdata/data/StormData.csv.bz2'
#download.file(data.url, destfile="storms.csv.bz2")
```

Now we can load the data we've downloaded:
```{r, cache=TRUE}
data.raw <- read.csv("storms.csv.bz2")
```

And consider some basic information about it:
```{r}
dim(data.raw)
summary(data.raw)
head(data.raw)
```

Create a dataset that is focused on health information:
```{r}
library(dplyr)
data.slim.health <- select(data.raw, BGN_DATE, STATE, EVTYPE, FATALITIES, INJURIES)

health <- group_by(data.slim.health, EVTYPE)
health.sum <- summarise(health, 
    count = n(),
    fatalities = sum(FATALITIES),
    injuries = sum(INJURIES),
    f_or_i = fatalities + injuries,
    i_per_event = injuries / count,
    f_per_event = fatalities / count,
    if_per_event = (injuries+fatalities)/count)
```

Create a dataset that is focused on economic information:
```{r}
# Limit the fields we need
data.slim.econ <- select(data.raw, BGN_DATE, STATE, EVTYPE, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

# Groupd by Event Type
econ <- group_by(data.slim.econ, EVTYPE)

# Crop Damage
conly <- filter(econ, CROPDMG > 0)

# We need to create actual dollar values
conly <- mutate(conly, cm = ifelse(CROPDMGEXP == "K", 1000, 
                            ifelse(CROPDMGEXP == "M", 1000000,
                            ifelse(CROPDMGEXP == "m", 1000000,
                            ifelse(CROPDMGEXP == "B", 1000000000, 1)))))
conly <- mutate(conly, cvalue = cm*CROPDMG)

# Property Damage
ponly <- filter(econ, PROPDMG > 0)

# We need to create actual dollar values
ponly <- mutate(ponly, pm = ifelse(PROPDMGEXP == "K", 1000, 
                            ifelse(PROPDMGEXP == "M", 1000000,
                            ifelse(PROPDMGEXP == "m", 1000000,
                            ifelse(PROPDMGEXP == "B", 1000000000, 1)))))
ponly <- mutate(ponly, pvalue = pm*PROPDMG)
```

## Results
### Across the United States, which types of events are most harmful with respect to population health?

We can see that Hurricane/Typhoon and Extreme Heat have traditionally caused the greatest impact on population health on a per event basis. In terms of total impact, Tornados have had the greatest affect on population health. The below chart is logrithmic due to the volume of people impacted by Tornados.
```{r, label="Results-Health"}
layout(matrix(c(1,1,2,2)))
tmp <- arrange(filter(health.sum, count > 10), desc(if_per_event))[1:10,]
par(las=2) # Make Label Perpendicular to Axis
par(mar=c(7,10,4,2)) # Adjust Y Margin
barplot(tmp$if_per_event, 
        names.arg=tmp$EVTYPE, 
        horiz=TRUE, 
        cex.names=0.6, 
        cex.axis=0.8,
        xlim=c(0,20),
        col="Dark Red",
        main="Injuries or Fatalities Per Event")

tmp <- arrange(filter(health.sum, count > 10), desc(f_or_i))[1:10,]
par(las=2) # Make Label Perpendicular to Axis
par(mar=c(7,10,4,2)) # Adjust Y Margin
barplot(tmp$f_or_i, 
        names.arg=tmp$EVTYPE, 
        horiz=TRUE, 
        cex.names=0.6, 
        cex.axis=0.8,
        log="x",
        xlim=c(1000,100000),
        col="Dark Red", 
        main="Total Injuries or Fatalities By Event Type")
```

### Across the United States, which types of events have the greatest economic impact?
In terms of crop damage, Drought has the greatest impact at nearly $14 Billion in total damages in the data set.
In terms of property damage, Flood has the greatest impact at over $144 Billion in total damages in the data set.
```{r, label="Results-Econ"}
layout(matrix(c(1,1,2,2)))
# Crop Damages
tmp <- arrange(summarise(conly, TCD=sum(cvalue)), desc(TCD))[1:10,]
par(las=2) # Make Label Perpendicular to Axis
par(mar=c(7,10,4,2)) # Adjust Y Margin
barplot(tmp$TCD, names.arg=tmp$EVTYPE,
        horiz=TRUE, cex.names = 0.6, cex.axis = 0.8,
        xlim=c(0, 1.5e+10), col="Dark Red", main="Crop Damage by Event Type")
tmp

# Property Damages
tmp <- arrange(summarise(ponly, TPD=sum(pvalue)), desc(TPD))[1:10,]
par(las=2) # Make Label Perpendicular to Axis
par(mar=c(7,10,4,2)) # Adjust Y Margin
barplot(tmp$TPD, names.arg=tmp$EVTYPE,
        horiz=TRUE, cex.names = 0.6, cex.axis = 0.8, log="x",
        xlim=c(1.0e+09, 1.0e+12), col="Dark Red", main="Log of Property Damage by Event Type")
tmp
```


## Data Exploration
### Health Exploration
Consider various information about the event types impact on population health, per event:
```{r}
# Events with the greatest number of injuries per event
arrange(health.sum, desc(i_per_event))[1:10,]

# Events with the greatest number of fatalities per event
arrange(health.sum, desc(f_per_event))[1:10,]

# Again, but only Event Types that have occured at least 10 times
arrange(filter(health.sum, count > 10), desc(i_per_event))[1:10,]
arrange(filter(health.sum, count > 10), desc(f_per_event))[1:10,]

# Let's consider both fatalities and injuries, as both impact population health
arrange(filter(health.sum, count > 10), desc(if_per_event))[1:10,]
```

Consider various information about the event types impact on population health, in total:
```{r}
# Events with the greatest number of injuries
arrange(health.sum, desc(injuries))[1:10,]

# Events with the greatest number of fatalities per event
arrange(health.sum, desc(fatalities))[1:10,]

# Let's consider both fatalities and injuries, as both impact population health
arrange(filter(health.sum, count > 10), desc(f_or_i))[1:10,]
```

Let's take a look at the various states:
```{r}
library(statebins)
states <- group_by(data.slim.health, STATE)
states.sum <- summarise(states, 
    count = n(),
    fatalities = sum(FATALITIES),
    injuries = sum(INJURIES),
    f_or_i = fatalities + injuries)

abvs <- c("AK", "AL", "AR", "AZ", "CA", "CO", "CT",
          "DC", "DE", "FL", "GA", "HI", "IA", "ID",
          "IL", "IN", "KS", "KY", "LA", "MA", "MD",
          "ME", "MI", "MN", "MO", "MS", "MT", "NC",
          "ND", "NE", "NH", "NJ", "NM", "NV", "NY",
          "OH", "OK", "OR", "PA", "RI", "SC", "SD", 
          "TN", "TX", "UT", "VA", "VT", "WA", "WI",
          "WV", "WY")
# Limited to 3 Figures in the final doc for the class
states.pure <- states.sum[states.sum$STATE %in% abvs,]
states.pure$state <- as.character(states.pure$STATE)
#statebins_continuous(as.data.frame(states.pure), brewer_pal="Purples",
#                     state_col="state", value_col="f_or_i")


```

Surprisingly Texas takes the worst of it.
```{r}
arrange(states.pure, desc(f_or_i))[1:10,]
```

### Economic Exploration
Let's calculate the total crop damage that is in the data set
```{r}
sum(arrange(summarise(conly, TCD=sum(cvalue)), desc(TCD))[1:10,2])
```

Now lets look at which states get hit the hardest by it.
```{r}
# Limited to 3 Figures in the final doc for the class
csts <- group_by(conly, STATE)
csts <- summarize(csts, count=n(), TCD=sum(cvalue))
csts.pure <- csts[csts$STATE %in% abvs,]
csts.pure$STATE <- as.character(csts.pure$STATE)
#statebins_continuous(as.data.frame(csts.pure), brewer_pal="Greens",
#                     state_col = "STATE", value_col = "TCD")

arrange(csts.pure, desc(TCD))[1:10,]
```

Let's calculate the total property damage that is in the data set
```{r}
sum(arrange(summarise(ponly, TPD=sum(pvalue)), desc(TPD))[1:10,2])
```

Now lets look at which states get hit the hardest by it.
```{r}
psts <- group_by(ponly, STATE)
psts <- summarize(psts, count=n(), TPD=sum(pvalue))
psts.pure <- psts[psts$STATE %in% abvs,]
psts.pure$STATE <- as.character(psts.pure$STATE)
statebins_continuous(as.data.frame(psts.pure), brewer_pal="Blues",
                     state_col = "STATE", value_col = "TPD")
arrange(psts.pure, desc(TPD))[1:10,]
```
